Computationally Efficient High-Dimensional Regularized Discriminant Analysis

### 2017-01-22 reports/paper-hdrda.Rnw
Timing comparison results discussion much improved.

# Build data frame to show relative timings in table
timing_ratios <- timings %>%
    group_by(p, Classifier) %>%
    summarize(median_time=median(time)) %>%
    tidyr::spread(Classifier, median_time) %>%
      mutate(ratio_time=RDA / HDRDA) %>%
      select(-HDRDA, -RDA)
@

@@

In this section, we demonstrate that the computational performance of the
training and model selection employed in the \emph{HDRDA} classifier is
substantially faster than that of the \emph{RDA} classifier on small-sample,
high-dimensional data sets. The relative difference in runtime between the two
classifiers drastically increases as $p$ increases. To
compare the two classifiers, we generated 25 observations from each of $K=4$
multivariate normal populations with mean vectors $\bm \mu_1 = -3 \cdot \bm
1_p$, $\bm \mu_2 = -\bm 1_p$, $\bm \mu_3 = \bm 1_p$, and $\bm \mu_4 = 3 \cdot
\bm 1_p$. We set the covariance matrix of each population to the $p \times p$
identity matrix. For each data set generated, we estimated the parameters $\lambda$ and
$\gamma$ for both classifiers using a grid of 5 equidistant candidate
values between 0 and 1, inclusively. We set $\alpha_k = 1 - \gamma$, $k = 1,
\ldots, K$, in the \emph{HDRDA} classifier. At each pair of
$\lambda$ and $\gamma$, we computed the 10-fold cross-validation error rate
\citep{Hastie:2008dt}. Then, we selected the optimal model that minimized the 10-fold
cross-validation error rate.

Here, our goal was to compare the computational runtime of the model selection
for both classifiers as a function of $p$. We compared the
runtime of both classifiers by increasing the number of features from $p = 500$
to $p = 5000$ in increments of 500. We generated 100 data sets for each value of
$p$ and computed the training and model selection runtime of both
classifiers. Our timing comparisons are based on our \emph{HDRDA} implementation
in the {\tt sparsediscrim} R package and the standard \emph{RDA} implementation
in the {\tt klaR} R package. All timing comparisons were conducted on an Amazon
Elastic Compute Cloud (EC2) {\tt c4.4xlarge} instance using version 3.3.1 of the
open-source statistical software {\tt R}. Our timing comparisons can be
reproduced with the code available in the
\url{https://github.com/ramhiser/paper-hdrda} repository.

@@

In Figure \ref{fig:timing-results}, we plotted the runtime of the model selections
for both the \emph{HDRDA} and \emph{RDA} classifiers as a function of $p$. In the left panel of Figure \ref{fig:timing-results}, we
observe that the \emph{HDRDA} classifier is substantially faster than the
\emph{RDA} classifier as $p$ increases. Figure \ref{fig:timing-results} suggests
that the usage of a matrix inverse and determinant in the {\tt klaR} R package's
discriminant function yielded training and model-selection timings that exceeded
linear growth in $p$. Because our implementation in the \emph{HDRDA} classifier
removes the inverse and determinants from the classifier, the \emph{HDRDA}
classifier is computationally more efficient than the \emph{RDA} classifier,
especially as $p$ increases. We contrast our implementation to the the poor
computational performance of the in the {\tt klaR} R package, which took
approximately 25 minutes on average to train the \emph{RDA} classifier on a data
set having 5,000 features. To quantify how much beyond linear growth in $p$, we
fit a regression line quadratic in $p$ to the \emph{RDA} runtimes and plotted
the fit in Figure \ref{fig:timing-results}. We displayed the
quadratic fit in Figure \ref{fig:timing-results} because it fit the timings best
for $p = 500$ to $p = 2000$. The fits for $2000 \le p \le 5000$ were similar.

Clearly, the model training and selection employed in \emph{HDRDA} is
substantially faster than that of \emph{RDA}. To investigate the relative timing
comparisons between the two classifiers, we considered the ratio of mean timings
of the \emph{RDA} classifier to the \emph{HDRDA} classifier for each value of
$p$. We employed nonparametric bootstrapping\footnote{TODO: Citation needed} to
estimate the mean ratio along with 95\% confidence intervals. In Figure
\ref{fig:timing-comparison-bootstrap}, the bootstrap sampling distributions for
the ratio of mean timings are given. First, we observe that the mean relative
timings increased as $p$ increased. For smaller dimensions, the relative
difference in computing was sizable with the average ratio of the mean timings
equal to \Sexpr{bootstrap_summary_p500$avg_ratio} for $p = 500$ and a 95\%
confidence interval of (\Sexpr{bootstrap_summary_p500$lower_ci},
\Sexpr{bootstrap_summary_p500$upper_ci}). Furthermore, the ratio of mean
computing times suggested that the \emph{RDA} classifier is impractical for
higher dimensions. For instance, when $p = 5000$ the ratio of mean computing
times increased to \Sexpr{bootstrap_summary_p5000$avg_ratio} with a 95\%
confidence interval of (\Sexpr{bootstrap_summary_p5000$lower_ci},
\Sexpr{bootstrap_summary_p5000$upper_ci}).

@@

We generated $K=3$ populations from \eqref{eq:contaminated-pdf} with $\bm
\Sigma_k$ given in \eqref{eq:block-sigma} and set the mean vector of class 1 to
$\bm \mu_1 = \bm 0_p$. Next, comparable to \cite{Guo:2007te}, the first 100
features of $\bm \mu_2$ were set to 1/2, while the rest were set to 0, i.e.,
$\bm \mu_2 = (\underbrace{1/2, \ldots, 1/2}_{100}, \underbrace{0, \ldots, 0}_{p
  - 100})$. For simplicity, we defined $\bm \mu_3 = -\bm \mu_2$. The three
populations differed in their mean vectors in the first 100 features
corresponding to the first block, and no difference in the means occurred in the
remaining blocks.

@@

We demonstrated that the \emph{HDRDA} implementation of
\eqref{eq:hdrda-decomposed} in the {\tt sparsediscrim} R package can be used in
practice with high-dimensional data sets. For instance, in our timing
comparisons we demonstrated that training and model selection could be employed
on data sets having 5,000 features in three seconds on average. We contrast our
implementation to the the poor computational performance of the in the {\tt
  klaR} R package, which took approximately 25 minutes on average to train the
\emph{RDA} classifier on a data set having 5,000 features. Given that the \emph{RDA}
classifier has been shown to have excellent performance in the high-dimensional
setting \citep{Webb:2011vu} but is limited by its computationally intense
model-selection procedure, our work enables the \emph{RDA} classifier to be
employed with small-sample, high-dimensional data in practice. This result is
reassuring because the \emph{RDA} classifier remains widely popular in the
literature. In fact, variants of the RDA classifier have been applied to
microarray data \citep{Ching:2012fu,Li:2012ev,Tai:2007bk,Guo:2007te}, facial
recognition
\citep{Ching:2012fu,Zhang:2010va,Dai:2007vd,Lu:2005hq,Pima:2004vw,Lu:2003we},
handwritten digit recognition \citep{Bouveyron:2007gx}, remote sensing
\citep{Tadjudin:1999fk}, seismic detection \citep{Anderson:2002kg}, and chemical
spectra \citep*{Wu:1996us,Aeberhard:1993fp}.

The dimension reduction that we have employed in this paper has reduced the
dimension to rank$(\widehat{\bm \Sigma}) = q$. An interesting extension of our
work would reduce the dimension $q$ further to a lower dimension $q_L < q$ using
some criterion proposed, perhaps similar to that employed in principal
components analysis. While unclear whether the classification performance would
improve via such a method, the efficiency of the training and model selection
would certainly improve. Moreover, if $q_L = 2$ or $3$ for low-dimensional graphical displays of high-dimensional data can
be obtained.

>>>>

# Build data frame to show relative timings in table
timing_ratios <- timings %>%
    group_by(p, Classifier) %>%
    summarize(median_time=median(time)) %>%
    tidyr::spread(Classifier, median_time) %>%
      mutate(ratio_time=RDA / HDRDA) %>%
      select(-HDRDA, -RDA)

timings_summary <- timings %>%
  group_by(Classifier, p) %>%
  summarize(mean_runtime=mean(time), sd_runtime=sd(time))

# RDA Mean Runtime in Minutes when p = 5000
rda_mean_time_p5000 <- timings_summary %>% filter(Classifier == "RDA", p == 5000)
rda_mean_time_p5000 <- rda_mean_time_p5000$mean_runtime / 60

# HDRDA Mean Runtime in Seconds when p = 5000
hdrda_mean_time_p5000 <- timings_summary %>% filter(Classifier == "HDRDA", p == 5000)
hdrda_mean_time_p5000 <- hdrda_mean_time_p5000$mean_runtime
@

@@

In this section, we demonstrate that the computational performance of the
model selection employed in the \emph{HDRDA} classifier is
substantially faster than that of the \emph{RDA} classifier on small-sample,
high-dimensional data sets. The relative difference in runtime between the two
classifiers drastically increases as $p$ increases. To
compare the two classifiers, we generated 25 observations from each of $K=4$
multivariate normal populations with mean vectors $\bm \mu_1 = -3 \cdot \bm
1_p$, $\bm \mu_2 = -\bm 1_p$, $\bm \mu_3 = \bm 1_p$, and $\bm \mu_4 = 3 \cdot
\bm 1_p$. We set the covariance matrix of each population to the $p \times p$
identity matrix. For each data set generated, we estimated the parameters $\lambda$ and
$\gamma$ for both classifiers using a grid of 5 equidistant candidate
values between 0 and 1, inclusively. We set $\alpha_k = 1 - \gamma$, $k = 1,
\ldots, K$, in the \emph{HDRDA} classifier. At each pair of
$\lambda$ and $\gamma$, we computed the 10-fold cross-validation error rate
\citep{Hastie:2008dt}. Then, we selected the model that minimized the 10-fold
cross-validation error rate.

We compared the
runtime of both classifiers by increasing the number of features from $p = 500$
to $p = 5000$ in increments of 500. We generated 100 data sets for each value of
$p$ and computed the training and model selection runtime of both
classifiers. Our timing comparisons are based on our \emph{HDRDA} implementation
in the {\tt sparsediscrim} R package and the standard \emph{RDA} implementation
in the {\tt klaR} R package. All timing comparisons were conducted on an Amazon
Elastic Compute Cloud (EC2) {\tt c4.4xlarge} instance using version 3.3.1 of the
open-source statistical software {\tt R}. Our timing comparisons can be
reproduced with the code available at
\url{https://github.com/ramhiser/paper-hdrda}.

@@

In Figure \ref{fig:timing-results}, we plotted the runtime of the model
selections for both the \emph{HDRDA} and \emph{RDA} classifiers as a function
of $p$. We observed that the \emph{HDRDA} classifier was substantially faster
than the \emph{RDA} classifier as $p$ increased. In the left panel of Figure
\ref{fig:timing-results}, we fit a quadratic regression line to the \emph{RDA}
runtimes and a simple linear regression model to the \emph{HDRDA} runtimes.
For improved understanding, in the right panel we repeated the same scatterplot
and linear fit with the timings restricted to the observed range of the
\emph{HDRDA} timings. Figure \ref{fig:timing-results} suggests that the usage
of a matrix inverse and determinant in the {\tt klaR} R package's discriminant
function yielded model-selection timings that exceeded linear growth in
$p$. Because the \emph{HDRDA} classifier removes inverse and determinants, it
was computationally more efficient than the \emph{RDA} classifier, especially
as $p$ increased. In fact, when $p = 5000$ the \emph{RDA} classifier required
\Sexpr{rda_mean_time_p5000} minutes on average to perform model selection,
while the \emph{HDRDA} classifier selected its optimal model in
\Sexpr{hdrda_mean_time_p5000} seconds on average. Clearly, the model selection
employed by the \emph{HDRDA} classifier is substantially faster than that of
the \emph{RDA} classifier.

We quantified the relative timing comparisons between the two classifiers by
calculating the ratio of mean timings of the \emph{RDA} classifier to the
\emph{HDRDA} classifier for each value of $p$. We employed nonparametric
bootstrapping to estimate the mean ratio along with 95\% confidence
intervals. In Figure \ref{fig:timing-comparison-bootstrap}, the bootstrap
sampling distributions for the ratio of mean timings are given. First, we
observe that the mean relative timings increased as $p$ increased. For smaller
dimensions, the relative difference in computing was sizable with the average
ratio of the mean timings equal to \Sexpr{bootstrap_summary_p500$avg_ratio} for
$p = 500$ and a 95\% confidence interval of
(\Sexpr{bootstrap_summary_p500$lower_ci},
\Sexpr{bootstrap_summary_p500$upper_ci}). Furthermore, the ratio of mean
computing times suggested that the \emph{RDA} classifier is impractical for
higher dimensions. For instance, when $p = 5000$ the ratio of mean computing
times increased to \Sexpr{bootstrap_summary_p5000$avg_ratio} with a 95\%
confidence interval of (\Sexpr{bootstrap_summary_p5000$lower_ci},
\Sexpr{bootstrap_summary_p5000$upper_ci}).

@@

We generated $K=3$ populations from \eqref{eq:contaminated-pdf} with $\bm
\Sigma_k$ given in \eqref{eq:block-sigma} and set the mean vector of class 1 to
$\bm \mu_1 = \bm 0_p$. Next, comparable to \cite{Guo:2007te}, the first 100
features of $\bm \mu_2$ were set to 1/2, while the rest were set to 0, i.e.,
$\bm \mu_2 = (\underbrace{1/2, \ldots, 1/2}_{100}, \underbrace{0, \ldots, 0}_{p
  - 100})$. For simplicity, we defined $\bm \mu_3 = -\bm \mu_2$. The 3
populations differed in their mean vectors in the first 100 features
corresponding to the first block, and no difference in the means occurred in the
remaining blocks.

@@

We demonstrated that the \emph{HDRDA} implementation of
\eqref{eq:hdrda-decomposed} in the {\tt sparsediscrim} R package can be used in
practice with high-dimensional data sets. For instance, in our timing
comparisons we demonstrated that training and model selection could be employed
on data sets having 5,000 features in 3 seconds on average. We contrast our
implementation to the the poor computational performance of the in the {\tt
  klaR} R package, which took approximately 25 minutes on average to train the
\emph{RDA} classifier on a data set having 5,000 features. Given that the \emph{RDA}
classifier has been shown to have excellent performance in the high-dimensional
setting \citep{Webb:2011vu} but is limited by its computationally intense
model-selection procedure, our work enables the \emph{RDA} classifier to be
employed with small-sample, high-dimensional data in practice. This result is
reassuring because the \emph{RDA} classifier remains widely popular in the
literature. In fact, variants of the RDA classifier have been applied to
microarray data \citep{Ching:2012fu,Li:2012ev,Tai:2007bk,Guo:2007te}, facial
recognition
\citep{Ching:2012fu,Zhang:2010va,Dai:2007vd,Lu:2005hq,Pima:2004vw,Lu:2003we},
handwritten digit recognition \citep{Bouveyron:2007gx}, remote sensing
\citep{Tadjudin:1999fk}, seismic detection \citep{Anderson:2002kg}, and chemical
spectra \citep*{Wu:1996us,Aeberhard:1993fp}.

The dimension reduction employed in this paper has reduced the
dimension to rank$(\widehat{\bm \Sigma}) = q$. An interesting extension of our
work would reduce the dimension $q$ further to a lower dimension $q_L < q$, perhaps using a criterion similar to that of principal
components analysis. While unclear whether the classification performance would
improve via such a method, the efficiency of the training and model selection
would certainly improve. Moreover, if $q_L = 2$ or $3$, low-dimensional graphical displays of high-dimensional data could
be obtained.
